---
title: "Using Steam's basic usage statistics to see which games have staying power"
date: 2020-02-19
output: html_notebook
---

# Introduction

A long while ago I saw [a presentation by Eric Seufert on using AppAnnie to do product development decisions and market research](http://www.slideshare.net/EricSeufert/using-free-app-annie-data-to-optimize-your-next-game) (for free). The idea is to take the daily iOS Top Grossing chart positions of games and then do a linear regression and find games that have a positive coefficient over time. It's really rough but gives some indication about games' lifecycles. One downside is that you only get the chart position when ideally you would want to know the daily player count or daily revenue (although you can get estimates of these through [AppAnnie's Market Intelligence](https://www.appannie.com/tours/market-data-intelligence/)).

Fortunately Steam has similar (and better) data available at [Steam & Game Stats page](http://store.steampowered.com/stats/). This page shows the current and peak players for top 100 games by current players. Ideally I would want to get the top 100 games by *peak* player count (or, daily active user (DAU) figure). Also, ideally, it would be nice to be able to browse this data over time instead of checking [Internet Archive](https://archive.org/help/wayback_api.php) and setting up a script to scrape the page every day but... (This data is something [SteamCharts](https://steamcharts.com) has but they don't have an API or anything).

```{r setup, include = FALSE}
library(tidyverse)
library(rvest)
library(lubridate)
library(arrow)
library(magrittr)
```

```{r import, eval=FALSE}
filenames <- list.files("steamstats", "\\.html$", full.names = TRUE) %>%
  str_subset("20190912", negate = TRUE) # %>% head(10)
# 20190912 is broken

read_steamstats <- function(fn) {
  html_data <- read_html(fn)
  df <- tibble(
    peak = html_data %>% 
      html_nodes("div#detailStats table tr.player_count_row") %>%
      html_nodes("td:nth-of-type(2) span") %>% 
      html_text() %>% 
      str_replace_all(",", "") %>% 
      strtoi(),
    game = html_data %>% 
      html_nodes("div#detailStats table tr.player_count_row") %>%
      html_nodes("td:nth-of-type(4)") %>% 
      html_text() %>% 
      str_trim() %>% 
      str_replace_all("(â¢|™|Â®|®)", ""),
    game_id = html_data %>% 
      html_nodes("div#detailStats table tr.player_count_row") %>% 
      html_nodes("td:nth-of-type(4) a") %>% html_attr("href") %>% 
      str_extract("(?<=/app/)\\d+") %>% str_trim(),
    date = ymd(str_extract(fn, "\\d{8}"))
  )
}

steam_stats <- lapply(filenames, read_steamstats) %>% 
  bind_rows()
```

```{r write cache, eval=FALSE, include=FALSE}
write_feather(steam_stats, "steam_stats.feather")
```

```{r read cache, include=FALSE}
steam_stats <- read_feather("steam_stats.feather")
```

# Data cleaning & data limitations

Just to remind the reader, this is just data from Steam, so this data gives no indication of these games' performance on consoles or on other launchers on PC (like Uplay or Epic Games). We are also only looking at games that managed to get at least once to top 100 list. There are tens of thousands of games on Steam, and we are limited to looking at barely `r length(unique(steam_stats$game_id))` of them.

```{r}
# remove Steam SDKs
steam_stats <- steam_stats %>%
  filter(!game %in% c("Source SDK Base 2013 Multiplayer", "Source SDK Base 2007"))
```

For some reason, sometimes for two annual franchises, Football Manager and Call of Duty, the game ID is missing. Fortunately, these can be found from the data.

```{r}
# entries missing a game_id
steam_stats %>% 
  filter(is.na(game_id)) %>% 
  group_by(game) %>% 
  summarise(missing = n())
```

For additional fun, Call of Duty: WWII - Multiplayer has two IDs in the dataset in addition to missing an ID (once). Other observations that have the same name, and have multiple IDs are:

```{r}
# game names that have multiple game id's.
steam_stats %>% 
  filter(!is.na(game_id)) %>% 
  group_by(game) %>% 
  summarise(game_id_count = length(unique(game_id)), game_ids = paste(unique(game_id), collapse = ", "), last_game_id = last(game_id, order_by = date)) %>%
  filter(game_id_count > 1)
```

We can look at the date ranges of these different game IDs and see that it looks like in most cases the game jumped to a new ID (no overlap) - except for Call of Duty: WWII Multiplayer, where the `476620` only blips for a few days. For our analysis, it doesn't really matter which ID is used (although it's likely that only the last one is active on Steam).

```{r}
# game names with multiple ids and the id's time spans
steam_stats %>% 
  filter(!is.na(game_id)) %>% 
  group_by(game) %>% 
  summarise(game_id_count = length(unique(game_id))) %>% 
  filter(game_id_count > 1) %>% 
  select(-game_id_count) %>% 
  left_join(steam_stats %>% 
              filter(!is.na(game_id)) %>% 
              group_by(game, game_id) %>% 
              summarise(min_date = min(date), max_date = max(date)), 
            by = "game") %>% 
  arrange(game, min_date)
```

Name changes are more common, usually these are weird punctuation changes (Elite: Dangerous -\> Elite Dangerous) or in case of Asian games, addition of English name. Or, in the case of for example H1Z1, evolution of the game. This suggests that charting the life cycle of a game can get really complicated.

```{r}
# game id's with multiple names
steam_stats %>% 
  filter(!is.na(game_id)) %>% 
  group_by(game_id) %>% 
  summarise(game_name_count = length(unique(game)), game_names = paste(unique(game), collapse = ", ")) %>% 
  filter(game_name_count > 1) %>% 
  arrange(desc(game_name_count))
```

```{r}
# summary table
steam_stats %>% 
  group_by(game_id) %>% 
  summarise(min_peak = min(peak), max_peak = max(peak), mean_peak = mean(peak), min_date = min(date), max_date = max(date), n = n(), game_name_count = length(unique(game)), game_names = paste(unique(game), collapse = ", "), ) %>%
  arrange(desc(max_peak))
```

```{r}
# fix data issues like in Python
# unify game_ids to the last seen game_id for game
# FIXME: this probably can be done simpler, fe. with grouping and dplyr::fill
unify_id_map <- steam_stats %>% 
  filter(!is.na(game_id)) %>% 
  group_by(game) %>% 
  summarise(game_id_count = length(unique(game_id)), last_game_id = last(game_id, order_by = date)) %>% 
  filter(game_id_count > 1) %>% 
  select(-game_id_count)
steam_stats <- steam_stats %>% 
  left_join(unify_id_map, by="game") %>% 
  mutate(game_id = if_else(is.na(last_game_id), game_id, last_game_id)) %>%
  select(-last_game_id) # fails if unify_id_map is empty
```

```{r}
# missing game_ids
steam_stats <- steam_stats %>% 
  group_by(game) %>% 
  mutate(game_id = ifelse(is.na(game_id), first(game_id), game_id)) %>% 
  ungroup()
```

```{r}
# multiple game names per id
steam_stats <- steam_stats %>% 
  group_by(game_id) %>% 
  mutate(game = last(game, order_by = date)) %>% 
  ungroup()
```

After this normalization, I have a dataset from `r format(min(steam_stats$date), format = "%B %Y")` to `r format(max(steam_stats$date), format = "%B %Y")` with (almost) daily snapshots of the Stats page from mid-August 2016 onward. There are some remaining data issues, especially with some annual franchises like Football Manager 201x - the player base quickly migrates to the next installation and the old game is removed from the Steam storefront. These installments could be grouped to one game, but where's the limit? Should Assassin's Creed games be considered as one franchise as well? Some other, mainly F2P, franchises re-use the game id, or for example in the case of H1Z1 and ARK: Survival Evolved split into two games at some point.

```{r}
steam_stats %>% 
  filter(grepl("Football Manager", game)) %>% 
  ggplot(., aes(x = date, y = peak, color = game)) + geom_line()
steam_stats %>% 
  filter(grepl("Football Manager", game)) %>% 
  group_by(date) %>% 
  summarise(sum_peak = sum(peak)) %>% 
  ggplot(., aes(x = date, y = sum_peak)) + geom_line() + geom_smooth(method = 'lm')
```

# Modeling engagement and growth

Note that this analysis does not tell anything about a game's revenue performance - however, what it tells is what kind of games are prime candidates for service-based business models. Instead, looking at the data it's pretty easy to identify which games are (or act like) traditional AAA-games with little post-launch planning and which games can sustain and grow their audience over time - be it through marketing efforts or pure virality or a combination of these. A couple of examples of such games are Rainbow Six Siege and Warframe. Seufert's idea was to summarise this graph as a simple linear regression: we can use the trend line's slope as a summary of how well the game's player base is doing.

```{r}
steam_stats %>% 
  filter(game %in% c("Tom Clancy's Rainbow Six Siege", "Warframe")) %>% 
  ggplot(., aes(x = date, y = peak, color = game)) + geom_line(aes(color = game)) + theme(legend.position = "bottom") + geom_smooth(method = 'lm')
```

The idea is to do this to all games in the dataset and find the games that go to the up and the right, ie. have a positive trend line.

```{r}
# build multiple models from the data frame (using the thing from r4ds)
game_model <- function(df) {
  lm(peak ~ date, data = df)
}

by_game <- steam_stats %>% 
  group_by(game, game_id) %>% 
  nest() %>%
  mutate(model = map(data, game_model))

models <- by_game %>% 
  mutate(glance = map(model, broom::glance), coeffs = map(model, broom::tidy)) %>% 
  select(-c(data, model)) %>% 
  unnest(c(glance, coeffs), names_repair = "universal") %>% 
  filter(term == "date") %>% 
  select(c(game, game_id, r.squared, estimate, nobs))
models %>% arrange(desc(estimate))
```

Many games only a quick appearance in the Top 100, either by having a big launch with little follow-up or due to a free weekend, full game giveaway or an open beta. We can filter these out by only looking at games that have been on the top list for at least 90 (non-consecutive) days. It's also worth to note that this is not a list of top played games played on Steam, this list is sorted by how much we have seen these games' player counts increase over time.

```{r}
models %>% filter(nobs >= 90) %>% arrange(desc(estimate))
```

I have also added the R squared metric, because it could be argued that games that better fit the linear model are "better" in growth. We can plot these games on these two metrics:

```{r}
ggplot(data = models %>% filter(nobs >= 90 && estimate > 0), aes(x = r.squared, y = estimate)) + geom_point() + geom_label(aes(label = game)) + scale_y_log10()
```

Above is a chart of all the games that have been on top list on more than 90 occasions (\~days) and which have in general seen their player base growing since the launch of the game. Y-axis (in log) is the coefficient (\~avg. player growth/day) and X-axis is the r\^2.

# Data exploration

Interesting cases of game's growth are for example Grand Theft Auto V. We see a traditional launch spike, but followed with a solid growth thanks to the GTA Online feature. The mid-year and year end spikes are related to Steam's biannual sales when the game has been available for a steep discount. The game was given away free on Epic Games Store in May 2020.

```{r}
# - Show Grand Theft Auto V as a AAA that transforms into growth with Online
steam_stats %>% 
  filter(game == "Grand Theft Auto V") %>% 
  ggplot(., aes(x = date, y = peak)) + geom_line() + geom_smooth(method = 'lm')
```

Destiny 2 was released on Steam as a free-to-play title in October 2019. It's normal that the initial interest wanes off for this kind of a AAA title, but the crucial thing for service based game is if it can recover from the "death valley".

```{r}
# - Destiny 2
steam_stats %>% 
  filter(game == "Destiny 2") %>% 
  ggplot(., aes(x = date, y = peak)) + geom_line() + geom_smooth(method = 'lm')
```

Annual franchises are often at the bottom of the list, because we see strong initial interest (due to in no small part by massive marketing campaigns) but these games are still designed only to last until the next installment and hope that the population moves (and grows).

```{r}
steam_stats %>% filter(grepl("Call of Duty", game)) %>% ggplot(., aes(x = date, y = peak, color = game)) + geom_line()
steam_stats %>% filter(grepl("Assassin's Creed", game)) %>% ggplot(., aes(x = date, y = peak, color = game)) + geom_line()
steam_stats %>% filter(grepl("Sid Meier's Civilization", game)) %>% ggplot(., aes(x = date, y = peak, color = game)) + geom_line()
```

If we looked at these stats just few years ago, the top performer would have been PUBG. It was a massive beast during 2018 - 2019, and still is formidable but it looks like its best days are behind it.

```{r}
# - Show a fad (PBUG, Dota Underlods?)
steam_stats %>% filter(game == "PLAYERUNKNOWN'S BATTLEGROUNDS") %>% ggplot(., aes(x = date, y = peak)) + geom_line() + geom_smooth(method = 'lm')
```

The true value of this analysis is to find those "smaller" games that are still growing, year on year. Especially the stories of niche games like Euro Truck Simulator 2 are very interesting. Path of Exile seems to really depend on regular content updates that drive players back. This can be a dangerous content treadmill.

```{r}
steam_stats %>% filter(game == "Rust") %>% ggplot(., aes(x = date, y = peak)) + geom_line() + geom_smooth(method = 'lm')
steam_stats %>% filter(game == "Path of Exile") %>% ggplot(., aes(x = date, y = peak)) + geom_line() + geom_smooth(method = 'lm')
steam_stats %>% filter(game == "Paladins") %>% ggplot(., aes(x = date, y = peak)) + geom_line() + geom_smooth(method = 'lm')
steam_stats %>% filter(game == "Euro Truck Simulator 2") %>% ggplot(., aes(x = date, y = peak)) + geom_line() + geom_smooth(method = 'lm')
```

We can also try to find the most "product-like" launches, with huge intial spikes and eventual fall off the charts. This games are often single player experiences with little replayability. Like GTA V, these games usually benefit from new player acquisition spikes during Steam's winter and summer sales.

```{r}
# - Show a AAA (Odyssey, FC5)
steam_stats %>% filter(game == "Assassin's Creed Odyssey") %>% ggplot(., aes(x = date, y = peak)) + geom_line() + geom_smooth(method = 'lm')
```

# Data visualization (extraneous)

We can look at weekly play seasonality. Unsurprisingly, we see higher gaming activity over the weekends. (Note that the data is a snapshot at midnight UTC so US gamers Sunday spills over to Monday).

```{r}
siege <- steam_stats %>% 
  filter(game %in% c("Tom Clancy's Rainbow Six Siege"))
siege %>% 
  mutate(wday = wday(date, label = TRUE)) %>% 
  ggplot(., aes(wday, peak)) + geom_boxplot() + facet_wrap(~ year(date))
```

# Next steps

This here is really just the first part of the approach taken in Seufert's presentation. The next step would be to group these games and find what elements correlate with the staying power. This can become a lot of manual work unless, for example, Steam's metadata on genres and user tags are used to enrich the dataset.
