---
title: "Using Steam's basic usage statistics to see which games have staying power"
date: 2026-02-06
author: "Kari Silvennoinen"
format: 
   html:
     df-print: kable
---

```{r}
#| label: setup
#| output: false

library(tidyverse)
library(broom)
library(ggrepel)

Sys.setlocale("LC_TIME", "C") # For English month names
options("lubridate.week.start" = 1)

steam_stats <- readRDS("steam_stats.rds")
```

## Introduction

A long time ago I saw [a presentation by Eric Seufert on <cite>Using (Free!) App Annie data to optimize your next game</cite>](https://www.slideshare.net/slideshow/using-free-app-annie-data-to-optimize-your-next-game/64352925)[^1]. The idea was to do market research by taking the daily iOS Top Grossing chart positions of games and then do a linear regression of them and so filter down to games that have a positive coefficient over time. It's really rough but does give some indication about games' life cycles and potential. One downside is that you only get the chart position when ideally you would want to know the daily player count or daily revenue (although you can get estimates of these if you pay for them).

[^1]: App Annie is now part of SensorTower, and this same analysis could be done with any source that gives you historical Top Grossing positions. App Annie just provided that information for free, although not for long after this presentation was shared.

Similar analysis can also be done on Steam, using the data available at [Steam & Game Stats page](http://store.steampowered.com/stats/). This page shows the current and peak players for top 100 games by current or daily players. The daily players is a more recent addition, back when I did my analysis below this option wasn't available. Instead of peak or current, ideally you would like to have a daily active user (<abbr>DAU</abbr>) figure, but either of these should work as a proxy for it. Also, ideally, it would be nice to be able to browse this data over time, but Steam doesn't provide this. This data is collected for example by [SteamCharts](https://steamcharts.com), but they don't have a public <abbr>API</abbr> for the historical data.

For your own purposes, you would need either have a script collect the data through [Steam Web <abbr>API</abbr>](https://steamcommunity.com/dev) like SteamCharts, and try to use services like [Internet Archive](https://archive.org/help/wayback_api.php) to get historical data and scrape the <abbr>HTML</abbr>.

## Data quality and limitations

In this analysis I'll use a data set ranging between `r format(min(steam_stats$date), format = "%B %Y")` and `r format(max(steam_stats$date), format = "%B %Y")`, with (almost) daily data from mid-August 2016 onwards (except around May 2018). The data collection method was a bit spotty in the beginning, because I was trying to get historical data through the Internet Archive's Wayback Machine.

```{r}
#| label: gaps in dates

dates <- unique(steam_stats$date)
expected <- seq(min(dates), max(dates), by="day")

cal <- tibble(
  date = expected,
  data_exists = if_else(expected %in% dates, "Yes", "No")
  ) |> 
  mutate(
  year = year(date),
  month = month(date, label = TRUE),
  wday = wday(date, label = TRUE, abbr = TRUE),
  isoweek = isoweek(date)
  ) |>
  group_by(year, month) |>
  mutate(week = if_else(
    month(date) == 1 & isoweek(date) >= 52,
    1,
    if_else(
      month(date) == 12 & isoweek(date) == 1,
      max(isoweek(date)) + 1,
      isoweek(date)
      )
    ) - min(
      if_else(month(date) == 12 & isoweek(date) == 1, 
              max(isoweek(date))+1, 
              isoweek(date)))+1
    ) |>
  ungroup()

ggplot(cal, aes(y = week, x = wday, fill = data_exists)) +
  geom_tile(color = "white") +
  coord_equal() +
  facet_grid(year ~ month) +
  scale_y_reverse(labels = NULL, breaks = NULL) +
  scale_x_discrete(labels = NULL, breaks = NULL) +
  labs(title = "Data availability", x = NULL, y = NULL, fill = "Data exists") +
  guides(fill = guide_legend(position = "bottom"))

```

Just to remind, this is data from Steam _only_, so it gives no indication of these games' performance on consoles or on other launchers on PC (like Epic Games or <abbr>GOG</abbr>). We are also only looking at games that managed to get at least _once_ to Top 100 list. There are tens of thousands of games on Steam, and we are limited to looking at barely `r length(unique(steam_stats$game_id))` of them.

Also note that this analysis does not tell anything about a game's revenue performance - we are looking at player counts, not revenue. However, what it tells is what kind of games are prime candidates for service-based business models. In fact, looking at the data it is pretty easy to identify which games are (or act like) traditional <abbr>AAA</abbr> games with little post-launch planning and which games can sustain and grow their audience over time - be it through marketing efforts or pure virality or a combination of these.

## Data prep and cleanup

Once we have parsed the data and have the raw data set, let's check the condition of the data we have. First we explore the uniqueness and existence of both Game IDs and game names and the cardinality between the two.

<cite>Football Manager</cite> together with <cite>Call of Duty</cite> have some entries missing the Game ID in the raw data. Fortunately, these can be found from elsewhere in the data.

```{r}
#| label: find missing game IDs
#| fig-cap: "Games with missing Game IDs"
#| fig-cap-location: top
steam_stats |>
  filter(is.na(game_id)) |>
  group_by(game) |>
  summarise(n = n()) |>
  arrange(-n) |>
  left_join(steam_stats |>
              filter(!is.na(game_id)) |>
              group_by(game_id) |>
              summarise(game=first(game)), by = "game") |>
  group_by(game) |>
  summarise(
    n = sum(n), 
    game_id = paste(unique(game_id), collapse = ", ")
    ) |>
  knitr::kable(col.names = c("Game", "N", "Game ID"), align = "lcc")
```


```{r}
steam_stats <- steam_stats |>
  group_by(game) |>
  mutate(game_id = if_else(is.na(game_id), min(game_id, na.rm = TRUE), game_id)) |>
  ungroup()
```

In the above table, you'll notice <cite>Call of Duty: WWII - Multiplayer</cite> matches two Game IDs. There's also some more non-uniqueness with game names in the data set.

```{r}
#| label: unify game ID
#| fig-cap: "Games with multiple Game IDs"
#| fig-cap-location: top

steam_stats |>
  group_by(game) |>
  summarise(
    count = length(unique(game_id)),
    ids = paste(unique(game_id), collapse = ", ")) |>
  filter(count > 1) |>
  arrange(-count) |>
  knitr::kable(col.names = c("Game", "Count", "Game IDs"), align = "lcc")
```

```{r}
steam_stats <- steam_stats |>
  group_by(game) |>
  mutate(game_id = min(game_id)) |>
  ungroup()
```

In addition there are also multiple names for the same Game ID. Name changes are somewhat common, usually these are weird punctuation changes (<cite>Elite: Dangerous</cite> -> <cite>Elite Dangerous</cite>) or in case of Asian games, addition of an English name. Or, in the case of for example <cite>H1Z1</cite>, evolution of the game. There was also some weirdness with addition and removal of trademark symbols, but those were filtered out already in the parsing of the raw data[^2]. This suggests that following the lifecycle of all games programmatically can get really complicated.

[^2]: Not shown here.

```{r}
#| label: unify game name
#| fig-cap: "Top 25 games with multiple names"
#| fig-cap-location: top

steam_stats |>
  filter(!is.na(game_id)) |>
  group_by(game_id) |>
  summarise(
    count = length(unique(game)),
    names = paste(unique(game), collapse = ", ")
    ) |>
  filter(count > 1) |> 
  arrange(-count) |> 
  head(25) |>
  knitr::kable(col.names = c("Game ID", "Count", "Game names"), align = "lcr")
```

```{r}
steam_stats <- steam_stats |>
  group_by(game_id) |>
  mutate(game = last(game)) |>
  ungroup()
```

With some annual franchises like <cite>Football Manager</cite> and 2K's <abbr>NBA</abbr> - the player base quickly migrates to the next installation a(nd sometimes like with the first franchise, the old game is removed from the Steam storefront). These installments could be grouped to one game for our purposes, but where should we set the limit? Should we do similar treatment for <cite>Call of Duty</cite> or just for annual sports franchises?

```{r}
#| label: combine Football Manager franchise
ggplot(steam_stats |> 
         select(peak_today, game, date) |> 
         filter(str_starts(game, "Football Manager")) |> 
         mutate(game = str_replace(game, "Football Manager", "FM")), 
       aes(x = date, y = peak_today, color = game)) +
  geom_line() +
  labs(title = "Daily peak CCU for Football Manager games", x = "Date", y = "Peak CCU last 24 hours", color = "Installment") +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale()))

steam_stats <- steam_stats |>
  mutate(game = if_else(str_starts(game, "Football Manager"), "Football Manager", game)) |>
  group_by(game) |>
  mutate(game_id = if_else(game == "Football Manager", min(game_id), game_id)) |>
  group_by(game, game_id, date) |>
  # NOTE: We assume distinct players across games, there can be overlap, we could also take max
  summarise(peak_today = sum(peak_today), current_players = sum(current_players), .groups = "drop") 

```

```{r}
#| label: combine NBA franchise
ggplot(steam_stats |> 
         select(peak_today, game, date) |> 
         filter(str_starts(game, "NBA 2K")), 
       aes(x = date, y = peak_today, color = game)) +
  geom_line() +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  labs(title = "Daily peak CCU for 2K's NBA games", x = "Date", y = "Peak CCU last 24 hours", color = "Installment") 

steam_stats <- steam_stats |>
  mutate(game = if_else(str_starts(game, "NBA 2K"), "NBA 2K", game)) |>
  group_by(game) |>
  mutate(game_id = if_else(game == "NBA 2K", min(game_id), game_id)) |>
  group_by(game, game_id, date) |>
  # NOTE: We assume distinct players across games, there can be overlap, we could also take max
  summarise(peak_today = sum(peak_today), current_players = sum(current_players), .groups = "drop") 
```

Now our data is in a good enough shape to continue with the actual modeling.

## Modeling engagement and growth

Seufert's idea was to summarise a game's activity graph as a simple linear regression: we can use the trend line's slope as a summary of how well the game's player base is doing. As an example, let's check what the Steam activity looks like for <cite>Warframe</cite> and <cite>Rainbow Six Siege</cite>[^3]. The idea is to do this for all games in the dataset and find the games that go to the up and the right, ie. have a positive trend line.

[^3]: These are poor examples in that they are multiplatform games that even on PC have their own launchers, so very much not Steam exclusive games.

```{r}
#| label: Siege and Warframe
steam_stats |> 
  filter(game %in% c("Tom Clancy's Rainbow Six Siege", "Warframe")) |>
  ggplot(aes(x = date, y = peak_today, color = game)) + 
  labs(title = "Daily peak CCU of Siege and Warframe", y = "Peak CCU last 24 hours", color = NULL) +
  geom_line(aes(color = game)) + 
  guides(color = guide_legend(position = "bottom")) +
  geom_smooth(method = 'lm') +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale()))
```

```{r}
#| label: run LM for the dataset
#| output: false
games <- steam_stats |>
  group_by(game) |>
  nest() |>
  mutate(
    first_date = map_vec(data, \(x) min(x$date)),
    model = map(data, \(x) lm(peak_today ~ date, data = x)),
    glance = map(model, broom::glance),
    tidy = map(model, broom::tidy)
    ) |>
  unnest(c(glance, tidy), names_repair = "unique") |>
  filter(term == "date") |>
  select(game, first_date, r.squared, nobs, estimate) |>
  rename(coeff = estimate) |>
  filter(!is.na(coeff)) |>
  ungroup()
```

```{r}
#| fig-cap: "Top 25 games by slope coefficient"
#| fig-subcap: "Only games with at least 90 observations"
#| fig-cap-location: top
games |>
  filter(nobs > 90) |>
  arrange(-coeff) |> 
  head(25) |>
  knitr::kable(col.names = c("Game", "First date", "$R^2$", "Observations", "Coefficient"), align = "lcc", digits = 2)
```

Many games only a quick appearance in the Top 100, either by having a big launch with little follow-up or due to a free weekend, full game giveaway or an open beta. We can filter these out by only looking at games that have been on the top list for at least 90 (non-consecutive) days. It's also worth to note that this is not a list of top played games played on Steam, this list is sorted by how much we have seen these games' player counts increase over time.

Above you'll notice <cite>Source <abbr>SDK</abbr> Base 2013 Multiplayer</cite>, it is used as the game identifier for some <cite><abbr>GTA</abbr> V</cite> mod. Likewise, a game called <cite>Spacewar</cite> might also pop up, it is a game included as a sample app in Steam <abbr>SDK</abbr> and often used as a placeholder by pirated games.  You might also wonder about <cite>Wallpaper Engine</cite>, and I'll let this [MIT Technology Review](https://www.technologyreview.com/2022/07/21/1056315/chinese-gamers-steam-wallpaper-porn/) article eplain that one. We'll filter these out.

```{r}
#| label: filter mods, porn and pirates
games <- games |>
  filter(!game %in% c("Source SDK Base 2013 Multiplayer", "Spacewar", "Wallpaper Engine"))
```

## Data exploration

Interesting cases of game's growth are for example <cite>Grand Theft Auto V</cite>. We see a traditional launch spike, but followed with a solid growth thanks to the <cite><abbr>GTA</abbr> Online</cite> feature. The mid-year and year end spikes are related to Steam's biannual sales when the game has been available for a steep discount. The game was given away free on Epic Games Store in May 2020, and this giveaway shows as increased player activity also on Steam.

```{r}
#| label: GTAV
steam_stats |>
  filter(game == "Grand Theft Auto V") |>
  ggplot(aes(x = date, y = peak_today, color = game)) +   
  labs(title = "Daily peak CCU of Grand Theft Auto V", y = "Peak CCU last 24 hours") +
  geom_line() + 
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  geom_smooth(method = 'lm', color = "black") +
  guides(color = "none")
```

<cite>Destiny 2</cite> was released on Steam as a free-to-play title in October 2019. It's normal that the initial interest wanes off for this kind of a <abbr>AAA</abbr> title, but the crucial thing for service based game is if it can recover from the "death valley".

```{r}
#| label: Destiny 2
steam_stats |>
  filter(game == "Destiny 2") |>
  ggplot(aes(x = date, y = peak_today, color = game)) + 
  labs(title = "Daily peak CCU of Destiny 2", y = "Peak CCU last 24 hours") +
  guides(color = "none") +
  geom_line() + 
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  geom_smooth(method = 'lm', color = "black")
```
Have we looked at the data early 2018, the top performer would have been <cite><abbr>PUBG</abbr></cite>. It was a massive beast during 2018 but it looks like <cite>Fortnite</cite> won the battle. <cite>Fornite</cite>, obviously, isn't on Steam so we can't compare it against <cite><abbr>PUBG</abbr></cite> here, but [charts comparing their viewership on Twitch](https://gamerant.com/fortnite-beat-pubg-twitch-stream-watch/) show <cite>Fortnite</cite> bypassing <cite><abbr>PUBG</abbr></cite> around February 2018, which is when we see <cite><abbr>PUBG</abbr></cite>'s activity on Steam start to crater.

```{r}
#| label: PUBG
steam_stats |>
  filter(game == "PUBG: BATTLEGROUNDS") |>
  ggplot(aes(x = date, y = peak_today, color = game)) + 
  labs(title = "Daily peak CCU of PUBG: BATTLEGROUNDS", y = "Peak CCU last 24 hours") +
  guides(color = guide_legend(position = "bottom")) +
  geom_line() + 
  geom_smooth(method = 'lm') +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  geom_line(aes(color = "PUBG until 2019"), data = steam_stats |> filter(game == "PUBG: BATTLEGROUNDS", date < as.Date("2019-01-01"))) +
  geom_smooth(aes(color = "PUBG until 2019"), data = steam_stats |> filter(game == "PUBG: BATTLEGROUNDS", date < as.Date("2019-01-01")), method = 'lm')

```
In our analysis we are looking at all the data, but we could limit the outcome window of games and only look at first <var>X</var> months after launch as well. The way we are doing our analysis here is a bit sensitive to the point of time we are looking at and biased toward new games that have yet to reach their plateau.

The true value of this analysis is to find those "smaller" games that are still growing, year on year. Especially the stories of niche games like <cite>Euro Truck Simulator 2</cite> are very interesting. The higher variance in <cite>Path of Exile</cite> could indicate it seems to really depend on regular content updates to drive players back. This can be a dangerous content treadmill. <cite><abbr>ETS2</abbr></cite> regularly adds new regions as <abbr>DLC</abbr>s, but it user activity seems to be growing in a more stable fashion.

```{r}
#| label: POE and ETS2
steam_stats |> 
  filter(game %in% c("Euro Truck Simulator 2", "Path of Exile")) |>
  ggplot(aes(x = date, y = peak_today, color  = game)) + 
  labs(title = "Daily peak CCU of Path of Exile and ETS2", y = "Peak CCU last 24 hours", color = NULL) +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  guides(color = guide_legend(position = "bottom")) +
  geom_line() + 
  geom_smooth(method = 'lm')
```

We can also look at "product-like" launches, with huge intial spikes and eventual fall off the charts - what I would call traditional games as products. This games are often single player experiences with little replayability. These games usually benefit from new player acquisition spikes during Steam's winter and summer sales.

```{r}
#| label: ACO
steam_stats |> 
  filter(game == "Assassin's Creed Odyssey") |>
  ggplot(aes(x = date, y = peak_today, color = game)) + 
  labs(title = "Daily peak CCU of Assassin's Creed Odyssey", y = "Peak CCU last 24 hours") +
  guides(color = "none") +
  geom_line() + 
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  geom_smooth(method = 'lm', color = "black")
```

## Clustering the games

In Seufert's presentation, he wanted to map the "opportunities", or games with an upward trajectory, on a matrix with "Mass appeal / Niche appeal" and "Male / Female" axes. I'll chart the games a little bit differently, based on the elements that correlate with the idea of "staying power". Below we'll just cluster based on the linear regression coefficients and top chart data. We could further enrich the data with furhter metadata from Steam, for example publisher-set genres and user-defined tags that could also be used as features for clustering or other segmentation.

```{r}
#| label: chart coeff vs. r^2
#| warning: false 
ggplot(data = games |> filter(coeff > 0, nobs > 90),
       aes(x = r.squared, y = coeff, label = game)) +
  labs(
    title = "Top games by slope that have more than 90 observations", 
    subtitle = "Games in top-right have grown most and most consistently",
    x = "R^2", 
    y = "Coefficient (log scale)") +
  geom_point() +
  scale_y_log10() +
  geom_text_repel(max.overlaps = 9, size = 3)
```
Above is a chart of all the games that have been on top list on more than 90 occasions (~days) and which have in general seen their player base growing since the launch of the game. Y-axis (in logarithmic scale) is the coefficient (~avg. player growth/day) and X-axis is the $R^2$.

We can try to quickly group these games a bit by clustering. The attributes used in clustering are the slope, days in top list and the $R^2$. We'll also this time include all the games, not just those that have a positive trend to also identify games that have more product-like activity curves.

```{r}
#| label: clustering
#| warning: false
#| fig-cap: "Cluster centers"
#| fig-cap-location: top

set.seed(42)

cluster_games <- games |> 
  filter(nobs > 90 & !is.na(coeff)) |>
  mutate(
    age = as.numeric(max(steam_stats$date) - first_date),
    logcoeff = if_else(coeff > 0, log10(coeff), -log10(-coeff))
    )

games_matrix <- cluster_games |> 
  select(-c(game, first_date, coeff)) |> 
  as.matrix() |> 
  scale()

rownames(games_matrix) <- cluster_games |> 
  pull(game)

km <- kmeans(games_matrix, centers = 4, nstart = 25)

# rescale cluster centers
km$centers |> 
  sweep(2, attr(games_matrix, "scaled:scale"), "*") |> 
  sweep(2, attr(games_matrix, "scaled:center"), "+") |> 
  as_tibble() |>
  knitr::kable(col.names = c("$R^2$", "Observations", "Age", "Log of Coefficient"), align = "c", digits = c(2, 0, 0, 4), row.names = TRUE)
  
```

We can make rough explanations for the clusters,

1. Games that do not have a strong explanatory power
2. Product-like games with a downward trend
3. Games with strong explanatory power, some of the opportunities can be found here
4. Service-like games that have upward trend and are mainstays on the top charts

```{r}
#| label: plot clusters
#| warning: false

cluster_games <- cluster_games |> 
  mutate(cluster = km$cluster)

ggplot(
  data = cluster_games, 
  aes(x = r.squared, y = logcoeff, label = game, color = as.factor(cluster))
  ) +
  labs(
    title = "Clustering of games",
    subtitle = "Games that have more than 90 observations", 
    x = "$R^2$", 
    color = "Cluster", 
    y = "Coefficient (log scale)"
  ) +
  geom_point() +
  geom_text_repel(max.overlaps = 12, size = 3)
```

## Conclusion

This is just a quick clustering, and very likely far from optimal. We still want to focus on the top-right corner of this map to find the winners and opportunities. We might want to segment the data based on the age of the game and as mentioned before, limit the outcome window of the games. We are also working with left-censored data, some of the big games had already plateaued by the start date of our data set, like <cite>Dota 2</cite>.

This analysis can be used as a quick way to get a better sense of the games on Steam's top charts and what their trajectories look like. It can't be overstated that this analysis is by necessity backward looking and it might be late to try to join the hype train as the market trends might change fast.
