---
title: "Using Steam's basic usage statistics to see which games have staying power"
date: 2026-02-06
author: "Kari Silvennoinen"
format: html
---

```{r}
#| label: setup
#| output: false
library(tidyverse)
library(broom)
library(ggrepel)

Sys.setlocale("LC_TIME", "C") # For English month names
options("lubridate.week.start" = 1)

steam_stats <- readRDS("steam_stats.rds")
```

## Introduction

A long time ago I saw [a presentation by Eric Seufert on <cite>Using (Free!) App Annie data to optimize your next game</cite>](https://www.slideshare.net/slideshow/using-free-app-annie-data-to-optimize-your-next-game/64352925)[^1]. The idea was to take the daily iOS Top Grossing chart positions of games and then do a linear regression of them and find games that have a positive coefficient over time. It's really rough but does give some indication about games' life cycles. One downside is that you only get the chart position when ideally you would want to know the daily player count or daily revenue.(although you can get estimates of these if you pay for them).

[^1]: App Annie is now part of SensorTower, and this same analysis could be done with any source that gives you historical Top Grossing positions. App Annie just provided that information for free, although not for long after this presentation was shared.

Similar analysis can also be done on Steam, using the data available at [Steam & Game Stats page](http://store.steampowered.com/stats/). This page shows the current and peak players for top 100 games by current or daily players. The daily players is a more recent addition, back when I did my analysis below this option wasn't available. Instead of peak or current, ideally you would like to have a daily active user (DAU) figure, but either of these should work as a proxy for it. Also, ideally, it would be nice to be able to browse this data over time, but Steam doesn't provide this. This data is collected for example by [SteamCharts](https://steamcharts.com), but they don't have a public API for the historical data.

For your own purposes, you would need either have a script collect the data through [Steam Web <abbr>API</abbr>](https://steamcommunity.com/dev) like SteamCharts, and try to use services like [Internet Archive](https://archive.org/help/wayback_api.php) to get historical data and scrape the <abbr>HTML</abbr>.

## Data quality and limitations

In this analysis I'll use a data set ranging between `r format(min(steam_stats$date), format = "%B %Y")` and `r format(max(steam_stats$date), format = "%B %Y")`, with (almost) daily data from mid-August 2016 onwards (except around May 2018).

```{r}
#| label: gaps in dates

dates <- unique(steam_stats$date)
expected <- seq(min(dates), max(dates), by="day")

cal <- tibble(
  date = expected,
  data_exists = if_else(expected %in% dates, "Yes", "No")
  ) |> 
  mutate(
  year = year(date),
  month = month(date, label = TRUE),
  wday = wday(date, label = TRUE, abbr = TRUE),
  isoweek = isoweek(date)
  ) |>
  group_by(year, month) |>
  mutate(week = if_else(
    month(date) == 1 & isoweek(date) >= 52,
    1,
    if_else(
      month(date) == 12 & isoweek(date) == 1,
      max(isoweek(date)) + 1,
      isoweek(date)
      )
    ) - min(
      if_else(month(date) == 12 & isoweek(date) == 1, 
              max(isoweek(date))+1, 
              isoweek(date)))+1
    ) |>
  ungroup()

ggplot(cal, aes(y = week, x = wday, fill = data_exists)) +
  geom_tile(color = "white") +
  coord_equal() +
  facet_grid(year ~ month) +
  scale_y_reverse(labels = NULL, breaks = NULL) +
  scale_x_discrete(labels = NULL, breaks = NULL) +
  labs(title = "Data availability", x = NULL, y = NULL, fill = "Data exists") +
  guides(fill = guide_legend(position = "bottom"))

```

Just to remind, this is data from Steam _only_ , so it gives no indication of these games' performance on consoles or on other launchers on PC (like Epic Games or GOG). We are also only looking at games that managed to get at least _once_ to Top 100 list. There are tens of thousands of games on Steam, and we are limited to looking at barely `r length(unique(steam_stats$game_id))` of them.

Also note that this analysis does not tell anything about a game's revenue performance - we are looking at player counts, not revenue. However, what it tells is what kind of games are prime candidates for service-based business models. In fact, looking at the data it is pretty easy to identify which games are (or act like) traditional AAA-games with little post-launch planning and which games can sustain and grow their audience over time - be it through marketing efforts or pure virality or a combination of these.

## Data cleanup

Exploring the data, some data issues pop up. First we explore the uniqueness and existence of both Game IDs and game names and the cardinality between the two.

Football Manager together with Call of Duty shave ome entries missing the Game ID in the raw data. Fortunately, these can be found from elsewhere in the data.

```{r}
#| label: find missing game IDs
(steam_stats |>
  filter(is.na(game_id)) |>
  group_by(game) |>
  summarise(n = n()) |>
  arrange(-n) |>
  left_join(steam_stats |>
              filter(!is.na(game_id)) |>
              group_by(game_id) |>
              summarise(game=first(game)), by = "game")) |>
  group_by(game) |>
  summarise(
    n = sum(n), 
    game_id = paste(unique(game_id), collapse = ", ")
    )

steam_stats <- steam_stats |>
  group_by(game) |>
  mutate(game_id = if_else(is.na(game_id), min(game_id, na.rm = TRUE), game_id)) |>
  ungroup()
```

In the above table, you'll notice <cite>Call of Duty: WWII - Multiplayer</cite> matches two Game IDs. There's also some more non-uniqueness with game names in the data set.

```{r}
#| label: unify game ID
(steam_stats |>
  group_by(game) |>
  summarise(
    count = length(unique(game_id)),
    ids = paste(unique(game_id), collapse = ", ")) |>
  filter(count > 1)) |>
  arrange(-count)

steam_stats <- steam_stats |>
  group_by(game) |>
  mutate(game_id = min(game_id)) |>
  ungroup()

```

In addition there are also multiple names for the same Game ID. Name changes are somewhat common, usually these are weird punctuation changes (Elite: Dangerous -> Elite Dangerous) or in case of Asian games, addition of an English name. Or, in the case of for example H1Z1, evolution of the game. There was also some weirdness with addition and removal of trademark symbols, but those were filtered out already in the parsing of the raw data[^2]. This suggests that following the lifecycle of all games programmatically can get really complicated.

[^2]: Not shown here.

```{r}
#| label: unify game name
(steam_stats |>
  filter(!is.na(game_id)) |>
  group_by(game_id) |>
  summarise(
    count = length(unique(game)),
    names = paste(unique(game), collapse = ", ")
    ) |>
  filter(count > 1) |> arrange(-count))

steam_stats <- steam_stats |>
  group_by(game_id) |>
  mutate(game = last(game)) |>
  ungroup()
```

With some annual franchises like <cite>Football Manager</cite> and 2K's NBA - the player base quickly migrates to the next installation and the old game is removed from the Steam storefront. These installments could be grouped to one game, but where should we set the limit? Should we do similar treatment for <cite>Call of Duty</cite> or just for annual sports franchises?

```{r}
#| label: combine Football Manager franchise
fm <- steam_stats |> select(peak_today, game, date) |> filter(str_starts(game, "Football Manager")) |> mutate(game = str_replace(game, "Football Manager", "FM"))
ggplot(fm, aes(x = date, y = peak_today, color = game)) +
  geom_line() +
  labs(title = "Daily peak CCU for Football Manager games", x = "Date", y = "Peak number of simultaneous players", color = "Installment") +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale()))
 #+stat_summary(aes(color = "Total"), fun = "sum", geom="line", linetype = 2, show.legend = TRUE)

steam_stats <- steam_stats |>
  mutate(game = if_else(str_starts(game, "Football Manager"), "Football Manager", game)) |>
  group_by(game) |>
  mutate(game_id = if_else(game == "Football Manager", min(game_id), game_id)) |>
  group_by(game, game_id, date) |>
  summarise(peak_today = sum(peak_today), current_players = sum(current_players), .groups = "drop") # NOTE: We assume distinct players across games, there can be overlap, we could also take max

```
```{r}
#| label: combine NBA franchise
nba2k <- steam_stats |> select(peak_today, game, date) |> filter(str_starts(game, "NBA 2K"))
ggplot(nba2k, aes(x = date, y = peak_today, color = game)) +
  geom_line() +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  labs(title = "Daily peak CCU for 2K's NBA games", x = "Date", y = "Peak number of simultaneous players", color = "Installment") 
# + stat_summary(aes(color = "Total"), fun = "sum", geom="line", linetype = 2, show.legend = TRUE)

steam_stats <- steam_stats |>
  mutate(game = if_else(str_starts(game, "NBA 2K"), "NBA 2K", game)) |>
  group_by(game) |>
  mutate(game_id = if_else(game == "NBA 2K", min(game_id), game_id)) |>
  group_by(game, game_id, date) |>
  summarise(peak_today = sum(peak_today), current_players = sum(current_players), .groups = "drop") # NOTE: We assume distinct players across games, there can be overlap, we could also take max

```


## Modeling engagement and growth

Seufert's idea was to summarise a game's activity graph as a simple linear regression: we can use the trend line's slope as a summary of how well the game's player base is doing. Let's check what the Steam activity looks like for Warframe and Rainbow Six Siege[^3]. The idea is to do this to all games in the dataset and find the games that go to the up and the right, ie. have a positive trend line.

[^3]: These are poor examples in that they are multiplatform games that even on PC have their own launchers, so very much not Steam exclusive games.

```{r}
steam_stats |> 
  filter(game %in% c("Tom Clancy's Rainbow Six Siege", "Warframe")) |>
  ggplot(aes(x = date, y = peak_today, color = game)) + 
  geom_line(aes(color = game)) + 
  guides(color = guide_legend(position = "bottom")) +
  geom_smooth(method = 'lm') +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale()))
```
```{r}
games <- steam_stats |>
  group_by(game) |>
  nest() |>
  mutate(
    first_date = map_vec(data, \(x) min(x$date)),
    model = map(data, \(x) lm(peak_today ~ date, data = x)),
    glance = map(model, broom::glance),
    tidy = map(model, broom::tidy)
    ) |>
  unnest(glance, names_repair = "unique") |>
  unnest(tidy, names_repair = "unique") |>
  filter(term == "date") |>
  select(game, first_date, r.squared, nobs, estimate) |>
  rename(coeff = estimate) |>
  ungroup()

(top_games <- games |>
  filter(nobs > 90) |>
  arrange(-coeff))
```

Many games only a quick appearance in the Top 100, either by having a big launch with little follow-up or due to a free weekend, full game giveaway or an open beta. We can filter these out by only looking at games that have been on the top list for at least 90 (non-consecutive) days. It's also worth to note that this is not a list of top played games played on Steam, this list is sorted by how much we have seen these games' player counts increase over time.

Above you'll notice <cite>Source SDK Base 2013 Multiplayer</cite>, it is used as the game identifier for some <cite>GTA V</cite> mod. Likewise, a game called <cite>Spacewar</cite> might also pop up, it is a game included as a sample app in Steam SDK and often used as a placeholder by pirated games.  You might also wonder about <cite>Wallpaper Engine</cite>, and I'll let this [MIT Technology Review](https://www.technologyreview.com/2022/07/21/1056315/chinese-gamers-steam-wallpaper-porn/) article eplain that one. We'll filter these out.

```{r}
#| label: filter mods, porn and pirates
games <- games |>
  filter(!game %in% c("Source SDK Base 2013 Multiplayer", "Spacewar", "Wallpaper Engine"))
top_games <- top_games |>
  filter(!game %in% c("Source SDK Base 2013 Multiplayer", "Spacewar", "Wallpaper Engine"))
```

## Data exploration

Interesting cases of game's growth are for example Grand Theft Auto V. We see a traditional launch spike, but followed with a solid growth thanks to the GTA Online feature. The mid-year and year end spikes are related to Steam's biannual sales when the game has been available for a steep discount. The game was given away free on Epic Games Store in May 2020, and this giveaway shows as increased player activity also on Steam.

```{r}
#| label: GTAV
steam_stats |>
  filter(game == "Grand Theft Auto V") |>
  ggplot(aes(x = date, y = peak_today)) + 
  geom_line() + 
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  geom_smooth(method = 'lm')
```

Destiny 2 was released on Steam as a free-to-play title in October 2019. It's normal that the initial interest wanes off for this kind of a AAA title, but the crucial thing for service based game is if it can recover from the "death valley".

```{r}
#| label: Destiny 2
steam_stats |>
  filter(game == "Destiny 2") |>
  ggplot(aes(x = date, y = peak_today)) + 
  geom_line() + 
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  geom_smooth(method = 'lm')
```
Have we looked at the data early 2018, the top performer would have been PUBG. It was a massive beast during 2018 but it looks like Fortnite won the battle. Fornite, obviously, isn't on Steam so we can't compare it against PUBG here, but [charts comparing their viewership on Twitch](https://gamerant.com/fortnite-beat-pubg-twitch-stream-watch/) show Fortnite bypassing PUBG around February 2018, which is when we see PUBG's activity on Steam start to crater.

```{r}
#| label: PUBG
steam_stats |>
  filter(game == "PUBG: BATTLEGROUNDS") |>
  ggplot(aes(x = date, y = peak_today, color = game)) + 
  geom_line() + 
  geom_smooth(method = 'lm') +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  geom_line(aes(color = "PUBG"), data = steam_stats |> filter(game == "PUBG: BATTLEGROUNDS", date < as.Date("2019-01-01"))) +
  geom_smooth(aes(color = "PUBG"), data = steam_stats |> filter(game == "PUBG: BATTLEGROUNDS", date < as.Date("2019-01-01")), method = 'lm')

```

The true value of this analysis is to find those "smaller" games that are still growing, year on year. Especially the stories of niche games like Euro Truck Simulator 2 are very interesting. The higher variance in Path of Exile could indicate it seems to really depend on regular content updates to drive players back. This can be a dangerous content treadmill. ETS2 regularly adds new regions as DLCs, but it user activity seems to be growing in a more stable fashion.

```{r}
steam_stats |> 
  filter(game %in% c("Euro Truck Simulator 2", "Path of Exile")) |>
  ggplot(aes(x = date, y = peak_today, color  = game)) + 
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  guides(color = guide_legend(position = "bottom")) +
  geom_line() + 
  geom_smooth(method = 'lm')
```

We can also try to find "product-like" launches, with huge intial spikes and eventual fall off the charts. This games are often single player experiences with little replayability. Like GTA V, these games usually benefit from new player acquisition spikes during Steam's winter and summer sales.

```{r}
steam_stats |> 
  filter(game == "Assassin's Creed Odyssey") |>
  ggplot(aes(x = date, y = peak_today)) + 
  geom_line() + 
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  geom_smooth(method = 'lm')
```

## Clustering the games

In Seufert's presentation, he wanted to map the "opportunities", or games with an upward trajectory, on a matrix with "Mass appeal / Niche appeal" and "Male / Female" axes. I'll chart the games a little bit differently, based on the elements that correlate with the staying power. Below we'll just cluster based on the linear regression coefficients and top chart data. Steam offers further metadata, for example publisher-set genres and user-defined tags that could also be used as factors for clustering or other grouping.

```{r}
ggplot(data = top_games |> filter(coeff > 0), aes(x = r.squared, y = coeff, label = game)) +
  labs(title = "Top games by slope that have more than 90 observations", x = "R^2") +
  geom_point() +
  scale_y_log10() +
  geom_text_repel(max.overlaps = 9, size = 3)
```

Above is a chart of all the games that have been on top list on more than 90 occasions (~days) and which have in general seen their player base growing since the launch of the game. Y-axis (in log) is the coefficient (~avg. player growth/day) and X-axis is the r^2.

We can try to quickly group these games a bit by grouping the list behavior of these games. The attributes used in clustering are the slope, days in top list and the r^2.

```{r}
# FIXME: Lots of random stuff here, and somehow PUBG is weirdly clustered but the Python one didn't scale the features, probably add last_date_of_data - first_date to balance out nobs
# note about point of time, new games probably still on upward trajectory before plateauing, could limit window to X days

set.seed(42)

games_matrix <- games |> filter(nobs > 14 & !is.na(coeff)) |> select(-c(game, first_date)) |> mutate(coeff = if_else(coeff > 0, log10(coeff), -log10(-coeff))) |> as.matrix() |> scale()
rownames(games_matrix) <- games |> filter(nobs > 14 & !is.na(coeff)) |> pull(game)

# library(factoextra)
# fviz_nbclust(games_matrix, kmeans, method = "wss")

km <- kmeans(games_matrix, centers = 4, nstart = 25)

# fviz_cluster(km, data = games_matrix)

# Cluster centers
km$centers |> sweep(2, attr(games_matrix, "scaled:scale"), "*") |> sweep(2, attr(games_matrix, "scaled:center"), "+")

clustered_games <- games |> filter(nobs > 14 & !is.na(coeff)) |> mutate(cluster = km$cluster) |> mutate(logcoeff = if_else(coeff > 0, log10(coeff), -log10(-coeff))) |> filter(nobs > 90)

ggplot(data = clustered_games, aes(x = r.squared, y = logcoeff, label = game, color = as.factor(cluster))) +
  labs(title = "Games that have more than 90 observations", x = "R^2", color = "Cluster") +
  geom_point() +
  geom_text_repel(max.overlaps = 9, size = 3)

```

